### 2020-11-03 latex/vlp.tex
VLP proof updated, sending to Manfredi for comments.

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:vlp-hamiltonian}) is
   \begin{equation}\label{eqn:eavg-dot-pos}
      \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      ,
   \end{equation}
   and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg}^- = - \dot{E}_\text{avg}^+
      .
   \end{equation}
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The positive-time derivative (\ref{eqn:eavg-dot-pos}) is nonnegative since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}^+} &= \sign{-\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{-\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= \sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\geq 0 \text{ (by assumption)}
      .
   \end{align}
   Likewise, the negative-time derivative (\ref{eqn:eavg-dot-neg}) is never
   positive, so \(E_\text{avg}\) is a Lyapunov function and the negative-time system is
   stable \cite{lyapunov}.

   Let us briefly analyze the upright equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Take now the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq
         E_\text{avg}(\pi,0) \right\}
      .
   \] 
   From (\ref{eqn:vlp-proof-neg-inv}), all sublevel sets of \(E_\text{avg}\) are
   negatively invariant in positive-time. 
   Hence, the complement \(\left(E_{(\pi,0)}\right)^C = \Gamma \backslash E_{(\pi,0)}\) is
   positively invariant.
   If the system is initialized on \(\left(E_{(\pi,0)}\right)^C\), it cannot
   posibly converge to \((\pi,0)\);
   hence, \(\Pi_0^+ \subset E_{(\pi,0)}\).
   Similarly, positive invariance of \(E_{(\pi,0)}\) in negative-time allows us
   to show that \(\Pi_0^- \subset \left(E_{(\pi,0)}\right)^C \cup (\pi,0)\).

@@

   We cannot linearize the dynamics at the origin to prove it is a repeller,
   because \(\theta\) is not defined there.
   Instead, let \(D := \Gamma \backslash \Pi_0^-\).
   We will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\)
   \cite{krasovskii_lasalle}.
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume 
   \(x(0) \in K\backslash \Pi_0^+\) and \(x(0) \neq (0,0)\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant, and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which, by Poincar\'{e}-Bendixson, is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.

   Suppose the system has a closed orbit. It would necessarily be contained in a
   level set of \(E_\text{avg}\) (since this is a non-decreasing function and
   closed orbits are periodic).
   Thus, \(E_\text{avg}\) is constant along the orbit, which means the
   orbit is contained in one of the axes because that is the only region where
   \(\dot{E}_\text{avg}^+ = 0\).
   Since the only invariant subsets of the axes are equilibria, the system cannot
   contain closed orbits.

   Hence, the positive limit set of \(x(t)\) must be one of the equilibria. 
   The origin is a repeller and \(x(0) \neq (0,0)\), which means \(x(t)\) is
   converging to \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

>>>>

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:vlp-hamiltonian}) is
   \begin{equation}\label{eqn:eavg-dot-pos}
      \dot{E}_\text{avg}^+ = -\frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      ,
   \end{equation}
   and the derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \begin{equation}\label{eqn:eavg-dot-neg}
      \dot{E}_\text{avg}^- = - \dot{E}_\text{avg}^+
      .
   \end{equation}
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The positive-time derivative (\ref{eqn:eavg-dot-pos}) is nonnegative since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}^+} &= \sign{-\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{-\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= \sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\geq 0 \text{ (by assumption)}
      .
   \end{align}
   Using (\ref{eqn:eavg-dot-neg}) and (\ref{eqn:vlp-proof-neg-inv}) 
   we can determine that \(\sign{\dot{E}_\text{avg}^-} \leq 0\). 
   Thus, \(E_\text{avg}\) is a Lyapunov function for
   (\ref{eqn:negative-time-vlp}), which means the negative-time system is stable
   \cite{lyapunov}.

   Let us briefly analyze the upright equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Take now the sublevel set
   \[
      E_{(\pi,0)} := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq
         E_\text{avg}(\pi,0) \right\}
      .
   \] 
   From (\ref{eqn:vlp-proof-neg-inv}), all sublevel sets of \(E_\text{avg}\) are
   negatively invariant in positive-time, which means
   complements of sublevel sets are positively invariant.
   Hence, the system cannot possibly converge to \((\pi,0)\) if it is
   initialized on 
   \(\left(E_{(\pi,0)}\right)^C = \Gamma \backslash E_{(\pi,0)}\).
   This implies that \(\Pi_0^+ \subset E_{(\pi,0)}\).
   Similarly, the interior of \(E_{(\pi,0)}\) is positively
   invariant for (\ref{eqn:negative-time-vlp}), 
   which implies \(\Pi_0^- \subset \left(E_{(\pi,0)}\right)^C \cup (\pi,0)\).

@@

   We cannot linearize (\ref{eqn:vlp-hamiltonian}) at \((0,0)\) to prove 
   the origin is a repeller, because \(\theta(0,0)\) is not well-defined. 
   Instead, let \(D := \Gamma \backslash \Pi_0^-\).
   We will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\)
   \cite{krasovskii_lasalle}.
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   In particular, the origin is asymptotically stable for every initial
   condition in the interior of \(E_{(\pi,0)}\).
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume 
   \(x(0) \in K\backslash \left(\Pi_0^+ \cup \left\{(0,0)\right\}\right)\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let \(E_k\) be its
   sublevel set.

   Since \(E_k\) contains \(K\) and \(x(t') \in K\), it follows that 
   \(x(t') \in E_k\). 
   Furthermore, negative invariance of \(E_k\) means the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\).
   However, \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which, by Poincar\'{e}-Bendixson, is either an equilibrium or a
   closed orbit \cite{poincare_bendixson}.
   We will show this positive limit set must be an equilibrium and thereby
   reach our contradiction.

   Suppose the system had a closed orbit. 
   The orbit would necessarily be contained in a
   level set of \(E_\text{avg}\), since this is a non-decreasing function and
   closed orbits are periodic.
   Thus, \(E_\text{avg}\) is constant along the orbit, which means the
   orbit is an invariant subset of the set \(Z\) consisting of the two axes.
   However, the only invariant subsets of \(Z\) are equilibria, so 
   the system cannot contain closed orbits.

   The lack of closed orbits means the positive limit set of \(x(t)\) must 
   be one of the equilibria. 
   The origin is a repeller and \(x(0) \neq (0,0)\), which means \(x(t)\) is
   converging to \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).
